Feeding Dynamic Behavior into Optimization: Simulation-based Multi-objective Overtime Planning

### 2014-09-03 data/result/Article/Latex/Article.tex
.

Continuous-time simulation has been extensively used to model software projects, most notably by the seminal work of Abdel-Hamid and Madnick \cite{Abdel-Hamid:1991}, who presented a System Dynamics model of a software project following a waterfall life-cycle and comprising more than 270 equations. Later, this model was extended by Tvedt \cite{Tvedt:1996}, who added equations to describe a concurrent and incremental life-cycle model. Lin et al. added requirement volatility, milestones, and the possibility to discard certain activities comprising the schedule due to lack of resources \cite{Lin:1997}. Madachy \cite{Madachy:1996} used System Dynamics to describe the behavior of inspections on software projects. Recently, Cao et al. \cite{Cao:2010} have used simulation models to evaluate the effects of practices suggested by agile processes.

Optimization has also been combined with simulation for project planning purposes. Harman et al. \cite{harman:2009} presents a general framework for project planning optimization that uses a simulator to calculate fitness functions according to a project plan and a set of constraints for staff allocation. Antoniol et al. \cite{Antoniol:2004} combine a genetic algorithm with a queuing simulator to optimize staff allocation and work package ordering on software projects. The simulator calculates project makespan taking into consideration uncertainties about the effort required for each activity, rework and abandonment of work packages. Recently, Rodriguez et al. \cite{Rodriguez:2011} used System Dynamics models in conjunction with NSGAII to find ideal team size and schedule estimations for a software project.

@@

In this paper we introduced a formulation for the overtime-planning problem that uses a multi-objective search-based approach and takes into account both positive and negative aspects of overtime. Continuous simulation is used to mimic the behavior of a software project under different conditions of overtime allocation. We compared the proposed approach with practices from the industry and with a similar formulation without the negative effects on quality.

We observed that our approach performs significantly better (high effect-size) than the margarine and critical-path overtime-allocation strategies. However, the practice that suggests allocating overtime in the second half of the project's schedule is competitive with our approach and may lead to good overtime allocations, specially in large projects. This behavior confirms the ``gut-feeling" of practitioners who tend to allocate overtime in the last activities to compensate for earlier delays. Therefore, by combining search-based optimization and software project simulation we can help researchers to design experimental studies to investigate, evaluate, and provide scientific support for practices and assumptions used in the industry, converting intuition into evidence-based knowledge.

For future work, we consider creating new simulation scenarios to model other aspects of software projects, such as exhaustion due excessive overtime, learning curve, inspections, among others. We also intend to integrate the approach with project management tools, allowing further experimental analysis, particularly to engage humans with the tool and observe their reaction to the overtime allocations it suggests. We also believe that creating means to help visualizing the differences among the overtime allocations proposed by NSGAII is an important aspect towards adopting our approach in practice.

>>>>

Continuous-time simulation has been extensively used to model software projects, most notably by the seminal work of Abdel-Hamid and Madnick \cite{Abdel-Hamid:1991}, who presented a System Dynamics model of a software project following a waterfall life-cycle and comprising more than 270 equations. Later, this model was extended by Tvedt \cite{Tvedt:1996}, who added equations to describe a concurrent and incremental life-cycle model. Lin et al. \cite{Lin:1997} added requirement volatility, milestones, and the possibility to discard certain activities comprising the schedule due to lack of resources. Madachy \cite{Madachy:1996} used System Dynamics to describe the behavior of inspections on software projects. Recently, Cao et al. \cite{Cao:2010} used simulation models to evaluate the effects of practices suggested by agile processes.

Optimization has also been combined with simulation for project planning purposes. Ferruci et al. \cite{ferrucci:2014} present a survey of search-based optimization applied to project management. Harman et al. \cite{harman:2009} present a general framework for project planning optimization that uses a simulator to calculate fitness functions according to a project plan and a set of constraints for staff allocation. Antoniol et al. \cite{Antoniol:2004} combine a genetic algorithm with a queuing simulator to optimize staff allocation and work package ordering on software projects. The simulator calculates project makespan taking into consideration uncertainties about the effort required for each activity, rework, and abandonment of work packages. Recently, Rodriguez et al. \cite{Rodriguez:2011} used System Dynamics models in conjunction with NSGAII to find ideal team size and schedule estimations for a software project.

@@

In this paper we introduced a formulation for the overtime-planning problem that uses a multi-objective search-based approach and takes into account both positive and negative aspects of overtime. Continuous simulation is used to mimic the behavior of a software project under different conditions of overtime allocation. We compared the proposed approach with practices from the industry and with a similar formulation without the negative effects of overtime on product quality.

We observed that our approach performs significantly better (high effect-size) than the margarine and critical-path overtime management strategies. However, the practice that suggests allocating overtime in the second half of the project's schedule is competitive with our approach and may lead to good overtime allocations, specially in large projects. This behavior confirms the ``gut-feeling" of practitioners who tend to allocate overtime in the last activities to compensate for earlier delays. Therefore, by combining search-based optimization and project simulation we can help researchers to design experimental studies to investigate, evaluate, and provide scientific support for practices and assumptions used in the industry, converting intuition into evidence-based knowledge.

For future work, we consider creating new simulation scenarios to model other aspects of software projects, such as exhaustion due excessive overtime, learning curve, and inspections. We also intend to integrate the approach with project management tools, allowing further experimental analysis, particularly to engage humans with the tool and observe their reaction to the overtime allocations it suggests. We also believe that creating means to help visualizing the differences among overtime allocations proposed by our approach is an important aspect towards its adoption in practice.
